% !TEX root = notes.tex
\section{Conclusion and future work}
\label{sec:conclusion}
Higher order logic is sometimes criticized as being too expressive. Some problems however, are most easily expressed using higher order, often due to a large modular structure. In this paper we used graph mining as an example of a higher order problem and made a thorough analysis of the problem from the knowledge representation point of view. While techniques exist to express these higher order problems in first order logic, sometimes, explicitly specifying the additional structure HO exhibits allows systems to perform better. For example, in the case of graph mining, higher order logic preserves the local coherence of graphs, and the independence of homomorphisms for the different examples, a property that a higher order solver can exploit in order to raise efficiency. In its current state however, no technique combines the expressiveness of higher order with high performance solving techniques. While ProB has higher modeling power, its solver support is currently not yet sufficient. Likewise, while the solver behind IDP and ASP is better, it places large burdens on the user to encode higher order in a way that the system can work with it.

        Inspired by this case study, we propose higher-order language extensions for IDP and propose alternative ways to implement them in the solver. In particular, as shown in Section \ref{sec:performance}, the use of subsolvers seems promising and will be further explored together with the idea of Benders decomposition \citep{Benders}. The performance of the encodings in IDP or ASP can be considered as the ultimate target. 
